Hydrodynamics and rheology of adhesive vesicle doublets

### 2019-01-14 vesicle_code/docs/adhesion/adhesion.tex
changes in intro and misc by yny

% ****** Start of file apssamp.tex ******
%
%   This file is part of the APS files in the REVTeX 4.1 distribution.
%   Version 4.1r of REVTeX, August 2010
%
%   Copyright (c) 2009, 2010 The American Physical Society.
%
%   See the REVTeX 4 README file for restrictions and more information.
%
% TeX'ing this file requires that you have AMS-LaTeX 2.0 installed
% as well as the rest of the prerequisites for REVTeX 4.1
%
% See the REVTeX 4 README file
% It also requires running BibTeX. The commands are as follows:
%
%  1)  latex apssamp.tex
%  2)  bibtex apssamp
%  3)  latex apssamp.tex
%  4)  latex apssamp.tex
%
\documentclass[aps,prl,showpacs,amsmath,amssymb,superscriptaddress]{revtex4-1}

@@

\section{Introduction}
Vesicles (closed fluid-filled phospholipid bilayer membranes) have been widely utilized
as biological cell mimics in biophysics and material engineering \cite{}. 
%
The ever expanding applications have encouraged detailed experimental, theoretical and numerical investigations of vesicle hydrodynamics
in applied flow, electric and magnetic fields. Theoretical investigations often take the form of small-deformation analysis or spheroidal dynamics of a nearly-spherical vesicle (or nearly-circular in two dimensions) subjected to an applied flow or under an external forcing \cite{Barthes-BieselRallison1981_JFM,Misbah2006_PRL,Vlahovska2007_PRE,Finken2008_EPL,ZhangZahnTanLin2013_PoF,Nganguia2013_PRE}. 
On the other hand, various numerical methods have been developed for simulating the transient hydrodynamics of vesicle suspensions  \cite{BagchiJohoson2005_JBE,Biben2005_EJP,Veerapaneni2009_JCP,SeolHuKimLai2016_JCP}.  
%Experiments are performed to study vesicle hydrodynamics under various flowing conditions or forcing configurations \cite{MaderVitkova2006_EurPhysJE, Abkarian2008_SoftMatt,DahlNarsimhanGouveia2016_SoftMatt}. 
Collective vesicle hydrodynamics naturally is dictated by the individual vesicle-vesicle interactions;  for example, 
it was shown recently that vesicles may align to form chains along the direction of the externally applied DC electric field with ensuing consequences to suspension rheology \cite{wu2018pairwise}.

In numerical simulations of vesicle suspensions, it is often assumed that the lubrication layer between vesicle membranes acts to keep the 
membrane-membrane interaction, such as the van der Waals type interaction, negligible. However the amphiphilic natures of lipid molecules (a hydrophobic tail and a hydrophilic head with an electric dipole) complicate the interactions between a lipid bilayer membrane and a solid (such as a glass substrate or a nano particle) or another bilayer membrane \cite{EvansMetcalfe1984_BJ,Book_PhysicalBasisCellAdhesion,Book_IntermolecularSurfaceForces}.
Studies confirmed that such interaction potential inevitably consists of a long-range attraction component and a short-range repulsion component 
\cite{Book_IntermolecularSurfaceForces}. Such interactions leads to adhesion at the separation distance that minimizes the
interaction potential \cite{Book_IntermolecularSurfaceForces}, and are essential in many biomedical, biological and biophysical processes.
For example, vesicle adhesion is essential for membrane fusion/fission in endocytosis and exocytosis, the transport of small
vesicles through membrane surfaces.

@@

This paper is organized as follows.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations}
We consider suspensions of locally inextensible vesicles in an unbounded
two-dimensional viscous fluid.  For simplicity, we assume the fluid
viscosity both inside and outside the vesicles is constant, but our
method can be easily adjusted to account for a viscosity contrast.
Individual vesicles are denoted as $\gamma_j$, and they are
parameterized in arclength as $\xx_j(s,t)$.  The union of all vesicles
is denoted by $\gamma$.  Given a background velocity $\uu_\infty$, the
governing equations are
\begin{equation*}
\begin{aligned}
  \mu \Delta \uu = \nabla p, \quad &\xx \in \RR^2,
    &&\mbox{\em conservation of momentum} \\
  \nabla \cdot \uu = 0, \quad &\xx \in \RR^2, 
    &&\mbox{\em conservation of mass} \\
  \uu \rightarrow \uu_\infty, \quad &\|\xx\| \rightarrow \infty,
    &&\mbox{\em far-field condition} \\
  \uu(\xx,t) = \dot{\xx}, \quad &\xx \in \gamma,
    &&\mbox{\em velocity continuity} \\
  \xx_s \cdot \uu_s =0, \quad &\xx \in \gamma,
    &&\mbox{\em vesicle inextensibility} \\
  \llbracket T \rrbracket \nn = \xxi, \quad &\xx \in \gamma,
    &&\mbox{\em non-zero traction jump}
\end{aligned}
\end{equation*}
where $\xxi$ is the traction jump that depends on a bending, stretching,
and adhesion.  The bending term is $\BB \xx = -\kappa_b \xx_{ssss}$,
where $\kappa_b$ is the bending modulus, and this term comes from the
Helfrich energy.  For the numerical examples, we set $\kappa_b = 1$ and
use a Hamaker constants ranging from $0.1$ to $1$.  The stretching term
is $\TT \sigma  = (\sigma \xx_s)_s$, where $\sigma$ is the tension.  The
resistance to bending and stretching are standard assumptions of
vesicles.  In this work, we include an adhesive term, $\AA \xx$, that we
now describe.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Adhesion Model}
The adhesion model of Sukumaran and Seifert~\cite{suk-sei2001} considers adhesion
between a single three-dimensional vesicle and a solid wall at $z=0$.
\todo[inline]{Need a cartoon picture in here.  Some kind of version of
the one I had for my SIAM Life Sciences Talk}
The adhesion potential in their model is of the functional form in equation~\ref{eq:adhesion_potential} with $(m,n) = (4,2)$:
\begin{align}
\label{eq:adhesion_potential}
  \phi(z) = \sigma \left[ 
    \left(\frac{\delta}{z}\right)^m - \frac{m}{n} \left(\frac{\delta}{z}\right)^n \right],
\end{align}
where $z$ is the distance between the solid wall and the vesicle,
$\sigma$ is the Hamaker constant, and $\delta$ is the adhesion length
scale.  We generalize this model to an unbounded suspension of vesicles.
We assume that the adhesion force applies between all pairs of points on
different vesicles, so we are interested in Hamaker constants that are
appropriate for lipid-lipid interactions. In
Appendix~\ref{sec:appendixA}, the adhesive force at a point $\xx$ on
vesicle $j$ is shown to be
\begin{align}
  \AA\xx:=-\sigma m \delta^{n+1}\sum_{\substack{k=1 \\ k \neq j}}^M 
  \int_{\gamma_k} \frac{\xx - \yy}{\|\xx - \yy\|^{m+1}} 
  \left(\delta^{m-n} - \|\xx - \yy\|^{m-n} \right) ds_\yy.
  \label{eqn:adhesionForce}
\end{align}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
\todo[inline]{Concluding remarks}

@@

We now decompose the perturbation into normal and tangential components
as $\delta \xx_1 = \epsilon \yy(s) = \epsilon(u\tt + v\nn)$. The
perturbed arclength term, to leading order, is
\begin{align*}
  \|\tilde{\xx}'_1\| \approx 1 + \epsilon(u_s + \nu\kappa),
\end{align*}
where $\kappa$ is the curvature.  To leading order, inextensible
perturbations satisfy $u_s + \kappa v = 0$, so the arclength term of
$\gamma_1$ and $\tilde{\gamma}_1$ are identical to leading order.
Therefore,
\begin{align*}
  \delta U_1 &= \int_{\gamma_1} \int_{\gamma_2} \left(
  \phi(\|\xx_1 + \epsilon \yy  - \xx_2\|) - \phi(\|\xx_1 - \xx_2\|)
  \right) ds_{\xx_2} ds_{\xx_1} \\
  &\approx \epsilon \int_{\gamma_1} \int_{\gamma_2}
  \nabla \phi (\|\xx_1 - \xx_2\|) \cdot \yy ds_{\xx_2} ds_{\xx_1}
\end{align*}
and the adhesive force applied by vesicle 2 on vesicle 1 is
\begin{align}
 \int_{\gamma_2}\nabla \phi(\|\xx_1 - \xx_2\|)ds_{\xx_2} = -m \sigma\delta^{n+1}
   \int_{\gamma_k} \frac{\xx - \yy}{\|\xx - \yy\|^{m+1}} 
  \left(\delta^{m-n} - \|\xx - \yy\|^{m-n} \right) ds_\yy.
  \end{align}
When $(m,n) = (4,2)$, the above express becomes
\begin{align*}
  \int_{\gamma_2}\nabla \phi(\|\xx_1 - \xx_2\|)ds_{\xx_2} = 
  -4 \sigma \delta^2 \int_{\gamma_2}
  \frac{\xx_1 - \xx_2}{\|\xx_1 - \xx_2\|^6} 
  \left(\delta^2 - \|\xx_1 - \xx_2\|^2 \right) ds_{\xx_2}.
\end{align*}
A similar expression holds for the adhesive force applied by vesicle 1
on vesicle 2, and equation~\eqref{eqn:adhesionForce} gives the adhesive
force for a suspension of $M$ vesicles.

>>>>

% ****** Start of file apssamp.tex ******
%
%   This file is part of the APS files in the REVTeX 4.1 distribution.
%   Version 4.1r of REVTeX, August 2010
%
%   Copyright (c) 2009, 2010 The American Physical Society.
%
%   See the REVTeX 4 README file for restrictions and more information.
%
% TeX'ing this file requires that you have AMS-LaTeX 2.0 installed
% as well as the rest of the prerequisites for REVTeX 4.1
%
% See the REVTeX 4 README file
% It also requires running BibTeX. The commands are as follows:
%
%  1)  latex apssamp.tex
%  2)  bibtex apssamp
%  3)  latex apssamp.tex
%  4)  latex apssamp.tex
%
\documentclass[aps,prb,showpacs,amsmath,amssymb,superscriptaddress]{revtex4-1}

@@

\section{Introduction}
Vesicles (closed fluid-filled phospholipid bilayer membranes) have been widely utilized
as biological cell mimics in biophysics and material engineering \cite{}. 
%
The ever expanding applications have encouraged detailed experimental, theoretical and numerical investigations of vesicle hydrodynamics
in applied flow, electric and magnetic fields. Theoretical investigations often take the form of small-deformation analysis or spheroidal dynamics of a nearly-spherical vesicle (or nearly-circular in two dimensions) subjected to an applied flow or under an external forcing \cite{Barthes-BieselRallison1981_JFM,Misbah2006_PRL,Vlahovska2007_PRE,Finken2008_EPL,ZhangZahnTanLin2013_PoF,Nganguia2013_PRE}. 
On the other hand, various numerical methods have been developed for simulating the transient hydrodynamics of vesicle suspensions  \cite{BagchiJohoson2005_JBE,Biben2005_EJP,Veerapaneni2009_JCP,SeolHuKimLai2016_JCP}.  
%Experiments are performed to study vesicle hydrodynamics under various flowing conditions or forcing configurations \cite{MaderVitkova2006_EurPhysJE, Abkarian2008_SoftMatt,DahlNarsimhanGouveia2016_SoftMatt}.

Hydrodynamics of a single vesicle in Stokes flow has been extensively investigated. In a planar shear flow, the vesicle hydrodynamics is characterized by
excess area (reduced volume), viscosity contrast between interior and exterior fluids, and shear rate of the imposed far-field fluid flow. In addition a vesicle
with a rigid particle inside is also investigated as a biological mimic of a 
eukaryotic cell with a nucleus that occupies nearly half of the intracellular volume \cite{Veerapaneni2011_PRL}. Small-deformation analysis shows that
a vesicle tank-treads in a planar shear flow for low viscosity contrast and shear rate. At high viscosity contrast the tank-tread (TT) dynamics transitions to tumbling (TB) \cite{Misbah2006_PRL,Vlahovska2007_PRE},
and this leads to a transition in effective shear viscosity in the vesicle suspension \cite{Misbah2006_PRL,Vitkova2008_BJ} that is also validated
by direct numerical simulations \cite{GhigliottiBibenMisbah2010_JFM} and experiments \cite{KantslerSegreSteinberg2008_EPL,ZabuskySegreDeschamps2011_PoF}.
Between TT and TB vesicle hydrodynamics, a breathing (tremble) is also observed \cite{Misbah2006_PRL,KantslerSegreSteinberg2008_PRL,ZhaoShaqfeh2011_JFM,SpannZhaoShaqfeh2014_PoF} to affect the effective shear viscosity in a 
different way.
In an extensional flow (planar or uniaxial), vesicle shape dynamics
depends sensitively on the vesicle reduced volume  and the elastic capillary number \cite{KantslerSegreSteinberg2008_PRL,ZhaoShaqfeh2013_JFM,Narsimhan2014_JFM,DahlNarsimhanGouveia2016_SoftMatt}: Asymmetric shape and oscillatory undulation of the vesicle membrane are two examples of the complex vesicle hydrodynamics in an extensional flow.

Collective dynamics of vesicless naturally is dictated by the individual vesicle-vesicle interactions.
%
% for example, 
%it was shown recently that vesicles may align to form chains along the direction of the externally applied DC electric field with ensuing consequences to suspension rheology \cite{wu2018pairwise}.
%
In numerical simulations of vesicle suspensions, it is often assumed that the liquid film between vesicles acts to keep the
membrane-membrane interaction, such as the van der Waals type adhesive interaction, negligible.
However the amphiphilic natures of lipid molecules (a hydrophobic tail and a hydrophilic head with an electric dipole) complicate the interactions between a lipid bilayer membrane and 
another bilayer membrane \cite{EvansMetcalfe1984_BJ,Book_PhysicalBasisCellAdhesion,Book_IntermolecularSurfaceForces} or
a solid (such as a glass substrate or a nano particle).
Studies confirmed that such interaction potential inevitably consists of a long-range attraction component and a short-range repulsion component 
\cite{Book_IntermolecularSurfaceForces}, and the combination of the two gives rise to adhesion at the separation distance that minimizes the
interaction potential \cite{Book_IntermolecularSurfaceForces}. Such adhesive potentials are essential in many biomedical, biological and biophysical processes.
For example, vesicle adhesion is essential for membrane fusion/fission in endocytosis and exocytosis, the transport of small
vesicles through membrane surfaces.

@@

This paper is organized as follows. In \S~\ref{sec:ge} we formulate our model for vesicle hydrodynamics with adhesive interactions  between membranes of distinct vesicles. 
We simulate the adhesion process of two vesicles in a quiescent flow in \S~\ref{sec:qflow}, where we also present a simple model to estimate how long it takes for two nearby
vesicles to reach an equilibrium configuration under adhesion. In \S~\ref{sec:eflow} we numerically investigate the hydrodynamics of a vesicle doublet in a fluid trap where the
stagnation point is actively controlled to be at the center of the vesicle doublet. From these results we propose a novel application of the microfluidic fluid trap to probe the 
adhesion strength between lipid bilayer membranes in solutions. In \S~\ref{sec:sflow} we investigate how two vesicles may  form a doublet as they move
toward each other under a planar shear flow.  We map out the parameter regions for bound/unbound vesicles under a planar shear flow, and we also investigate the effects of
adhesive interactions on the rheological properties of a dilute suspension of vesicle doublets. 
Finally in \S~\ref{sec:conclusions} we discuss the implications of our results and point out a few
potential future directions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations \label{sec:ge}}
We consider suspensions of locally inextensible vesicles in an unbounded
two-dimensional viscous fluid.  For simplicity, we assume the fluid
viscosity both inside and outside the vesicles is constant, but our
method can be easily adjusted to account for a viscosity contrast.
Individual vesicles are denoted as $\gamma_j$, and they are
parameterized in arclength as $\xx_j(s,t)$.  The union of all vesicles
is denoted by $\gamma$.  Given a background velocity $\uu_\infty$, the
governing equations are
\begin{equation*}
\begin{aligned}
  \mu \Delta \uu = \nabla p, \quad &\xx \in \RR^2,
    &&\mbox{\em conservation of momentum} \\
  \nabla \cdot \uu = 0, \quad &\xx \in \RR^2, 
    &&\mbox{\em conservation of mass} \\
  \uu \rightarrow \uu_\infty, \quad &\|\xx\| \rightarrow \infty,
    &&\mbox{\em far-field condition} \\
  \uu(\xx,t) = \dot{\xx}, \quad &\xx \in \gamma,
    &&\mbox{\em velocity continuity} \\
  \xx_s \cdot \uu_s =0, \quad &\xx \in \gamma,
    &&\mbox{\em vesicle inextensibility} \\
  \llbracket T \rrbracket \nn = \xxi, \quad &\xx \in \gamma,
    &&\mbox{\em non-zero traction jump}
\end{aligned}
\end{equation*}
where $\xxi$ is the traction jump that depends on a bending, stretching,
and adhesion.  The bending term is $\BB \xx = -\kappa_b \xx_{ssss}$,
where $\kappa_b$ is the bending modulus, and this term comes from the
Helfrich energy.  For the numerical examples, we set $\kappa_b = 1$ and
use a Hamaker constants ranging from $0.1$ to $1$.  The stretching term
is $\TT \sigma  = (\sigma \xx_s)_s$, where $\sigma$ is the tension.  The
resistance to bending and stretching are standard assumptions of
vesicles.  In this work, we include an adhesive term, $\AA \xx$, that we
now describe.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Adhesion Model}
The adhesion model of Sukumaran and Seifert~\cite{suk-sei2001} considers adhesion
between a single three-dimensional vesicle and a solid wall at $z=0$.
\todo[inline]{Need a cartoon picture in here.  Some kind of version of
the one I had for my SIAM Life Sciences Talk}
The adhesion potential in their model is of the functional form in equation~\ref{eq:adhesion_potential} with $(m,n) = (4,2)$:
\begin{align}
\label{eq:adhesion_potential}
  \phi(z) = \sigma \left[ 
    \left(\frac{\delta}{z}\right)^m - \frac{m}{n} \left(\frac{\delta}{z}\right)^n \right],
\end{align}
where $z$ is the distance between the solid wall and the vesicle,
$\sigma$ is the Hamaker constant, and $\delta$ is the adhesion length
scale.  We generalize this model to an unbounded suspension of vesicles.
We assume that the adhesion force applies between all pairs of points on
different vesicles, so we are interested in Hamaker constants that are
appropriate for lipid-lipid interactions. In
Appendix~\ref{sec:appendixA}, the adhesive force at a point $\xx$ on
vesicle $j$ is shown to be
\begin{align}
  \AA\xx:=-\sigma m \delta^{n}\sum_{\substack{k=1 \\ k \neq j}}^M 
  \int_{\gamma_k} \frac{\xx - \yy}{\|\xx - \yy\|^{m+2}} 
  \left(\delta^{m-n} - \|\xx - \yy\|^{m-n} \right) ds_\yy.
  \label{eqn:adhesionForce}
\end{align}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions\label{sec:conclusions}}
\todo[inline]{Concluding remarks}

@@

We now decompose the perturbation into normal and tangential components
as $\delta \xx_1 = \epsilon \yy(s) = \epsilon(u\tt + v\nn)$. The
perturbed arclength term, to leading order, is
\begin{align*}
  \|\tilde{\xx}'_1\| \approx 1 + \epsilon(u_s + \nu\kappa),
\end{align*}
where $\kappa$ is the curvature.  To leading order, inextensible
perturbations satisfy $u_s + \kappa v = 0$, so the arclength term of
$\gamma_1$ and $\tilde{\gamma}_1$ are identical to leading order.
Therefore,
\begin{align*}
  \delta U_1 &= \int_{\gamma_1} \int_{\gamma_2} \left(
  \phi(\|\xx_1 + \epsilon \yy  - \xx_2\|) - \phi(\|\xx_1 - \xx_2\|)
  \right) ds_{\xx_2} ds_{\xx_1} \\
  &\approx \epsilon \int_{\gamma_1} \int_{\gamma_2}
  \nabla \phi (\|\xx_1 - \xx_2\|) \cdot \yy ds_{\xx_2} ds_{\xx_1}
\end{align*}
and the adhesive force applied by vesicle 2 on vesicle 1 is
\begin{align}
 \int_{\gamma_2}\nabla \phi(\|\xx_1 - \xx_2\|)ds_{\xx_2} = -m \sigma\delta^{n}
   \int_{\gamma_k} \frac{\xx - \yy}{\|\xx - \yy\|^{m+2}} 
  \left(\delta^{m-n} - \|\xx - \yy\|^{m-n} \right) ds_\yy.
  \end{align}
When $(m,n) = (4,2)$, the above express becomes
\begin{align*}
  \int_{\gamma_2}\nabla \phi(\|\xx_1 - \xx_2\|)ds_{\xx_2} = 
  -4 \sigma \delta^2 \int_{\gamma_2}
  \frac{\xx_1 - \xx_2}{\|\xx_1 - \xx_2\|^6} 
  \left(\delta^2 - \|\xx_1 - \xx_2\|^2 \right) ds_{\xx_2}.
\end{align*}
A similar expression holds for the adhesive force applied by vesicle 1
on vesicle 2, and equation~\eqref{eqn:adhesionForce} gives the adhesive
force for a suspension of $M$ vesicles.
